Gravitational Lens Recovery with \Glass: Illustrating the power of non-parametric lens modelling

### 2013-06-12 limits/ms.tex
More work done on the theory section and general structure. Slowly getting there.

@@

\title[\Glass]{Gravitational Lens Recovery with \Glass: Illustrating the power of free-form lens modelling}
\author{%
Jonathan P. Coles 
\and 
Justin I. Read
\and 
Prasenjit Saha 
}

@@

\begin{abstract}
{\bf JR. Does \Glass\ stand for something?}\\
We introduce a new free-form gravitational lens modelling tool \Glass\ that builds on an earlier code \PixeLens. Like \PixeLens, we use a grid of mass pixels to model the lens, searching through many models consistent with the data to marginalise over model uncertainties. However, \Glass\ improves on \PixeLens\ in a number of important respects: (i) it uses a significantly improved sampling strategy that is both faster and results in significantly smaller errors; (ii) it is built on a fully modular framework that can support different basis functions; (iii) non-linear constraints can be applied in post-processing (for example, filters to remove models with spurious additional images or models inconsistent with stellar dynamics data are already implemented); (iv) the stellar mass (if known) can be applied as a prior; and (v) adaptive resolution can be used in regions of interest. We apply \Glass\ to a suite of dynamically realistic mock N-body data to determine what quality of data -- from lensing, stellar kinematics and/or stellar masses -- are required to measure the circularly averaged mass profile of a lens and its shape. Our key findings are as follows: (i) for pure lens data, multiple sources with wide redshift separation give the strongest constraints as this breaks the well-known mass-sheet or steepness degeneracy; (ii) a single quad with time delays also performs well, giving a good recovery of both the mass profile and its shape; (iii) stellar masses -- for lenses where the stars dominate the central potential -- can also break the steepness degeneracy, giving a recovery almost as good as having time delay data or multiple source redshifts; (iv) stellar kinematics are similarly powerful if the data extend beyond the half light radius of the stars $r \gg r_{1/2}$ (required to integrate out the effect of the unknown velocity anisotropy, $\beta(r)$); and (v) if the lensing data already probe the mass profile over the region $r \sim r_{1/2}$, then stellar kinematic data can be used to probe $\beta(r)$ -- an interesting dynamic quantity in its own right. Where information on the mass distribution from lensing and other probes becomes redundant, this opens up the possibility of using strong lensing to constrain cosmological models. We will study this, and present the first results from \Glass\ applied to real data, in forthcoming papers.
\end{abstract}

\section{Introduction}\label{sec:intro}
{\bf JR. Lots of references missing here. Need to be added. But this should get us started.}\\

@@

The majority of lens modelling efforts to date have focussed on `parametric' modelling, where simple theoretically motived models are fit using maximum likelihood methods to the data \citep{2011A&ARv..19...47K, 1993A&A...273..367K, 2010GReGr..42.2151K}. This is useful in that a simple model provides valuable physical insight. However, as data quality improves and the questions we ask of these data become more refined, it is important to re-asses the impact of our model assumptions. `Non-parameteric' methods -- that assume many more parameters than the available data constraints -- are invaluable in this respect. In non-parametric mass modelling, the lens inversion problem is deliberately chosen to be under-constrained \citep{1997MNRAS.292..148S,2005MNRAS.360..477D,2010ApJ...723.1678C,2006MNRAS.367.1209L,2013arXiv1304.2393S}. Instead of finding one `best-fit' model, we must search for many models consistent with the data, exploring degeneracies between different but plausible\footnotemark\ mass models \citep{2000AJ....120.1654S,2006ApJ...653..936S,2008MNRAS.386..307L}. The key advantages of non-parametric modelling are that we can: (i) explore model degeneracies and marginalise over them; and (ii) we can determine what type and quality of data (including data from stellar population synthesis modelling, stellar kinematics, or other indepedent probes of the mass) are required to measure parameters of interest.

@@

In this paper, we present a new free-form or `non-parametric' strong lens modelling framework \Glass\ that builds on an earlier code \PixeLens\ \citep{1997MNRAS.292..148S,2008ApJ...679...17C}. Like \PixeLens, we use a grid of mass pixels to model the lens, searching through many models consistent with the data to marginalise over model uncertainties. However, \Glass\ improves on \PixeLens\ in a number of important respects:

@@

\section{Theoretical Background}\label{sec:theory}

\subsection{What can we learn from strong lensing data?}

@@

\subsection{Lensing}

The lensing equation
%
\begin{equation}
\vec\beta = \vec\theta - \frac{D_{LS}}{D_S}\vec\alpha(\vec\theta)
\label{lensing equation}
\end{equation}
%
maps an observed image position $\vec\theta$ to a source position $\vec\beta$
via the correction factor 
%
\begin{equation}
\vec\alpha(\vec\theta) = \frac{4GD_L}{c^2} \int \Sigma(\vec\theta')\frac{(\vec\theta - \vec\theta')}{\ |\vec\theta - \vec\theta'|^2}d^2\vec\theta'
\end{equation}
%
which accounts for the gravitational lensing effect of an intervening mass. For
our purposes we make use of the thin mass sheet approximation and assume that
the lensing mass is infinitely thin compared with the distance between the
source and the lens $D_{LS}$ and the lens and the observer $D_{L}$. These
distances are angular diameter distances where $D_* = (c/H_0)d_*$ and $d_*$ is
a dimensionless number that depends on cosmology.  The lens can then be thought
of as a projected surface density $\Sigma$ which diverts the path of a photon
instantaneously through the bending angle $\vec\alpha$.

The arrival (or travel) time $t$ of a photon depends not only on the geometric
path it takes, but also on the general relativistic gravitational time dilation
due to the lens at a redshift of $z_L$.  In its complete form, with all units
and cosmology included we can write the arrival time as
%
\begin{equation}
N^2ct(\vec\theta) = (1+z_L)\frac{D_{L}D_{S}}{D_{LS}}\frac12 |\vec\theta - \vec\beta|^2 - (1+z_L)\frac{4GD_{L}^2}{c^2}\int \Sigma(\vec\theta') \ln |\vec\theta-\vec\theta'| d^2\vec\theta'
\label{full arrival time}
\end{equation}
%
The non-standard factor of $D_L^2$ in the second term comes from the fact that $\Sigma$
has units of \Msun/lyr$^2$. We can clean this up by writing down a dimensionless time
delay 
%
\begin{equation}
\tau = \left[(1+z_L)D_{L}\right]^{-1}ctN^2 = \left[ (1+z_L) \frac{c}{H_0}d_L\right]^{-1}ctN^2 = \left[ (1+z_L) d_L\right]^{-1}\nu t
\label{tau}
\end{equation}
%
in terms of our previous definitions. If we
further define a dimensionless density
%
\begin{equation}
\kappa_\infty = \frac{4\pi G}{c^2}\frac{c}{H_0}d_L\Sigma
              = \frac{d_L}{\nu}\Sigma
\end{equation}
%
and a lensing potential
%
\begin{equation}
\psi(\vec\theta) = \frac1\pi \int \kappa_\infty(\vec\theta') \ln|\vec\theta - \vec\theta'| d^2\vec\theta'\
\label{lensing potential}
\end{equation}
%
we can express \eqnref{full arrival time} very compactly as
%
\begin{equation}
\tau(\vec\theta) = \frac12 \xi |\vec\theta-\vec\beta|^2 - \psi(\vec\theta)
\label{arrival time}
\end{equation}
%
where $\xi \equiv D_{S}/D_{LS}$. We explicitly write $\kappa_\infty$ to remind ourselves
that there is no source distance factor involved. This will be useful later when we consider
multiple sources.

If we think of $\tau$ as a surface then images will be observed where ever
$\nabla \tau = 0$. In other words, images obey Fermat's principle and appear at
extremums of the surface. Note that $\nabla \tau$ is just \eqnref{lensing equation} or
\begin{equation}
  \xi \vec\beta = \xi \vec\theta - \frac 1\pi\int \kappa_\infty\frac{(\vec\theta - \vec\theta')}{\ |\vec\theta - \vec\theta'|^2}d^2\vec\theta'
\end{equation}

\subsection{Degeneracies} %-----------------------------------------------------

%
%The steepness degeneracy arbitrarily rescales the arrival time surface. Image
%positions are left alone but the density profile will become shallower or
%steeper. This is particularly bad for inferring the Hubble constant, as the
%lens profile will tend to become steeper for larger values of $H_0$.
%
%Other degeneracies include ring degeneracies, where one is allowed to build
%up a density distribution as a series of mass rings; the presence of substructure;
%and the effects of merging galaxies. 
%
%Shape degeneracy---a twisting in the density.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Numerical Methods}\label{sec:glass}

@@

For this paper we will restrict ourselves to using the pixelated basis set that
has been employed in previous work and is implemented in \PixeLens. Here we
will briefly review the main ideas. The algorithm for generating models in
\Glass\ samples a convex polytope in a high dimensional space whose interior
points are solutions to the lensing equation and satisfy other physically
motivated linear priors (the sampling algorithm in \Glass\ is not the same as
employed in \PixeLens).  We therefore formulate all of our equations as
equations linear in the unknowns. We describe the density distribution $\kappa$
as a set of discrete grid cells or pixels $\kappa_i$ and rewrite the potential
(\eqnref{lensing potential}) as
%
\begin{equation}
  \psi(\vec\theta) = \sum_n \kappa_n Q_n(\vec\theta)
  \label{discrete potential}
\end{equation}
%
where the sum runs over all the pixels and $Q_n$ is the integral of the logarithm
over pixel $n$. The exact form for $Q$ is described in \appref{Q derivation}.
We can find the discretized lens equation by simply taking the gradient of the
above equations.

@@

\subsection{Raytracing}
\label{Raytracing}
\Glass\ can also determine the position of images and time delays from 
particle-based simulation output given a source position $\vec\beta$. This is
used to generate the lens configurations used in the parameter study.  The
particles are first projected onto a very high resolution grid representing the
lens plane. The centers $\vec\theta_i$ of each of the grid cells are mapped
back onto the source plane using \eqnref{lensing equation}. If the location on
the source plane $\vec\beta_i$ is within a user specified
$\eps_\mathrm{accept}$ of $\vec\beta$ then $\vec\theta_i$ is 
accepted and further refined using a root finding algorithm until the distance
to $\vec\beta$ is nearly zero. If multiple points converge to an
$\eps_\mathrm{root}$ of each other then only one point is taken.  Care must be
taken that the grid resolution is high enough that the resulting image position
error is below the equivalent observational error. Time delays are then
calculated in order from the arrival time at each image (\eqnref{tau}).

\section{Parameter study}

@@

\subsection{Mock galaxy data}

@@

Each of these configurations were modeled with and without time delays and with
and without a central image. We assumed for all our tests that the lensing mass
was radially symmetric. By construction, this is known to be true and this is
most often the case with real galaxies, unless there is an obvious observed
asymmetry. The central pixel was refined into a further nine pixels to capture
any steep rise in the profile. Two of the four mock galaxies have a steeply rising
inner profile.

@@

The main results from modeling these different configuration are shown black in
\figref{main results}. Each subplot corresponds to a different galaxy and
the vertical axis shows the range of quality.  The quality of a
model $\M_i$ by comparing the recovered density profile $\rho_\M(r)$ against
the profile of the mock galaxy $\rho_G(r)$.  In particular, the plots show the
RMS quantity
%
\begin{equation}
  \chi_i = 100 \times \sqrt{\frac{\int_r (\rho_{\M_i}(r) - \rho_G(r))^2}{\int_r \rho_G(r)^2}}
  %\chi_i = \frac{\sum_r (\rho(r)_{\M_i} - \rho(r)_G)^2}{\sum_r \rho(r)_G^2}
\end{equation}
%

@@

\section{Results} %--------------------------------------------------------------

%\subsection{Shape fitting}

%\subsection{Using principle component analysis to find new degeneracies}

\section{Discussion}

>>>>

\def\newblock{\hskip .11em plus .33em minus .07em}

@@

\title[\Glass]{Gravitational Lens Recovery with \Glass: Illustrating the power of non-parametric lens modelling}
\author{%
Jonathan P. Coles 
\and 
Justin I. Read
\and 
Prasenjit Saha 
}

@@

\begin{abstract}
{\bf JR. Does \Glass\ stand for something?}\\
We introduce a new non-parametric gravitational lens modelling tool \Glass\ that builds on an earlier code \PixeLens. Like \PixeLens, we use a grid of mass pixels to model the lens, searching through many models consistent with the data to marginalise over model uncertainties. However, \Glass\ improves on \PixeLens\ in a number of important respects: (i) it uses a significantly improved sampling strategy that is both faster and results in significantly smaller errors; (ii) it is built on a fully modular framework that can support different basis functions; (iii) non-linear constraints can be applied in post-processing (for example, filters to remove models with spurious additional images or models inconsistent with stellar dynamics data are already implemented); (iv) the stellar mass (if known) can be applied as a prior; and (v) adaptive resolution can be used in regions of interest. We apply \Glass\ to a suite of dynamically realistic mock N-body data to determine what quality of data -- from lensing, stellar kinematics and/or stellar masses -- are required to measure the circularly averaged mass profile of a lens and its shape. Our key findings are as follows: (i) for pure lens data, multiple sources with wide redshift separation give the strongest constraints as this breaks the well-known mass-sheet or steepness degeneracy; (ii) a single quad with time delays also performs well, giving a good recovery of both the mass profile and its shape; (iii) stellar masses -- for lenses where the stars dominate the central potential -- can also break the steepness degeneracy, giving a recovery almost as good as having time delay data or multiple source redshifts; (iv) stellar kinematics are similarly powerful if the data extend beyond the half light radius of the stars $r \gg r_{1/2}$ (required to integrate out the effect of the unknown velocity anisotropy, $\beta(r)$); and (v) if the lensing data already probe the mass profile over the region $r \sim r_{1/2}$, then stellar kinematic data can be used to probe $\beta(r)$ -- an interesting dynamic quantity in its own right. Where information on the mass distribution from lensing and other probes becomes redundant, this opens up the possibility of using strong lensing to constrain cosmological models. We will study this, and present the first results from \Glass\ applied to real data, in forthcoming papers.
\end{abstract}

\section{Introduction}\label{sec:intro} %-----------------------------------------------------

{\bf JR. Lots of references missing here. Need to be added. But this should get us started.}\\

@@

The majority of lens modelling efforts to date have focussed on `parametric' modelling, where simple theoretically motived models are fit using maximum likelihood methods to the data \citep{2011A&ARv..19...47K, 1993A&A...273..367K, 2010GReGr..42.2151K}. This is useful in that a simple model provides valuable physical insight. However, as data quality improves and the questions we ask of these data become more refined, it is important to re-asses the impact of our model assumptions. `Non-parameteric' methods -- that assume many more parameters than the available data constraints -- are invaluable in this respect. In non-parametric mass modelling, the lens inversion problem is deliberately chosen to be under-constrained \citep{1997MNRAS.292..148S,2005MNRAS.360..477D,2010ApJ...723.1678C,2006MNRAS.367.1209L,2013arXiv1304.2393S}. Instead of finding one `best-fit' model, we must then search for many models consistent with the data, exploring degeneracies between different but plausible\footnotemark\ mass models \citep{2000AJ....120.1654S,2006ApJ...653..936S,2008MNRAS.386..307L}. The key advantages of non-parametric modelling are that we can: (i) explore model degeneracies and marginalise over them; and (ii) we can determine what type and quality of data (including data from stellar population synthesis modelling, stellar kinematics, or other indepedent probes of the mass) are required to measure parameters of interest.

@@

In this paper, we present a new non-parametric strong lens modelling framework \Glass\ that builds on an earlier code \PixeLens\ \citep{1997MNRAS.292..148S,2008ApJ...679...17C}. Like \PixeLens, we use a grid of mass pixels to model the lens, searching through many models consistent with the data to marginalise over model uncertainties. However, \Glass\ improves on \PixeLens\ in a number of important respects:

@@

\section{Theoretical Background}\label{sec:theory} %-----------------------------------------------------

\subsection{Lensing basics}\label{sec:lensing_basic} %-----------------------------------------------------

The lens equation:
%
\begin{equation}
\vec\beta = \vec\theta - \frac{D_{LS}}{D_S}\vec\alpha(\vec\theta)
\label{eqn:lens_equation}
\end{equation}
%
maps an observed image position $\vec\theta$ to a source position $\vec\beta$
via the correction factor:
%
\begin{equation}
\vec\alpha(\vec\theta) = \frac{4GD_L}{c^2} \int \Sigma(\vec\theta')\frac{(\vec\theta - \vec\theta')}{\ |\vec\theta - \vec\theta'|^2}d^2\vec\theta'
\end{equation}
%
which accounts for the gravitational lensing effect of an intervening mass \citep[e.g.][]{1992grle.book.....S}. For
our purposes we make use of the thin mass sheet approximation and assume that
the lensing mass is infinitely thin compared with the distance between the
source and the lens $D_{LS}$ and the lens and the observer $D_{L}$. These are angular diameter distances where $D_* = (c/H_0)d_*$ and $d_*$ is
a dimensionless number that depends on cosmology.  The lens can then be thought
of as a projected surface density $\Sigma$ which diverts the path of a photon
instantaneously through the bending angle $\vec\alpha$.

One way to understand the lens equation is via Fermat's principle. We can think of light as travelling only 
along extremum paths; lensed images only occur at the extrema of a virtual photon {\it arrival time surface} 
\citep{1986ApJ...310..568B}. The arrival (or travel) time $t$ of a virtual photon depends not only on the geometric
path it takes, but also on the general relativistic gravitational time dilation
due to the lens at a redshift of $z_L$.  In its complete form, with all units
and cosmology included we can write the arrival time as \citep{1986ApJ...310..568B}:
%
\begin{eqnarray}
N^2ct(\vec\theta) & = & (1+z_L)\frac{D_{L}D_{S}}{D_{LS}}\frac12 |\vec\theta - \vec\beta|^2 \nonumber \\
& & - (1+z_L)\frac{4GD_{L}^2}{c^2}\int \Sigma(\vec\theta') \ln |\vec\theta-\vec\theta'| d^2\vec\theta'
\label{eqn:full_arrival_time}
\end{eqnarray}
%
where the factor of $D_L^2$ in the second term comes from the fact that $\Sigma$
has units of \Msun/lyr$^2$. We can clean this up by writing down a dimensionless time
delay:
%
\begin{eqnarray}
\tau & = & \left[(1+z_L)D_{L}\right]^{-1}ctN^2 \nonumber \\
& = & \left[ (1+z_L) \frac{c}{H_0}d_L\right]^{-1}ctN^2 \nonumber \\
& = & \left[ (1+z_L) d_L\right]^{-1}\nu t
\label{tau}
\end{eqnarray}
%
in terms of our previous definitions. If we
further define a dimensionless density:
%
\begin{equation}
\kappa_\infty = \frac{4\pi G}{c^2}\frac{c}{H_0}d_L\Sigma
              = \frac{d_L}{\nu}\Sigma
\end{equation}
%
and a lensing potential:
%
\begin{equation}
\psi(\vec\theta) = \frac1\pi \int \kappa_\infty(\vec\theta') \ln|\vec\theta - \vec\theta'| d^2\vec\theta'\
\label{lensing potential}
\end{equation}
%
we can express \eqnref{eqn:full_arrival_time} very compactly as:
%
\begin{equation}
\tau(\vec\theta) = \frac12 \xi |\vec\theta-\vec\beta|^2 - \psi(\vec\theta)
\label{arrival time}
\end{equation}
%
where $\xi \equiv D_{S}/D_{LS}$. We explicitly write $\kappa_\infty$ to remind ourselves
that there is no source distance factor involved. This will be useful later when we consider
multiple sources.

If we think of $\tau$ as a surface then images will be observed wherever
$\nabla \tau = 0$. In other words, images obey Fermat's principle and appear at
extremums of the surface. Note that $\nabla \tau$ is just equation \ref{eqn:lens_equation} or:
\begin{equation}
  \xi \vec\beta = \xi \vec\theta - \frac 1\pi\int \kappa_\infty\frac{(\vec\theta - \vec\theta')}{\ |\vec\theta - \vec\theta'|^2}d^2\vec\theta'
\end{equation}

\subsection{Degeneracies}\label{sec:degen} %-----------------------------------------------------

The trouble with solving the lens equation for the mass distribution is that the solutions are non-unique and degenerate -- many different solutions can give a statistically good reproduction of the data. To see why this is, it is instructive to simplify equation \ref{eqn:lens_equation} for a circularly symmetric lens:

@@

\section{Numerical Methods}\label{sec:glass}

@@

For this paper we will restrict ourselves to using the pixelated basis set that
has been employed in previous work and is implemented in \PixeLens. Here we
will briefly review the main ideas. The algorithm for generating models in
\Glass\ samples a convex polytope in a high dimensional space whose interior
points are solutions to the lens equation and satisfy other physically
motivated linear priors (the sampling algorithm in \Glass\ is not the same as
employed in \PixeLens).  We therefore formulate all of our equations as
equations linear in the unknowns. We describe the density distribution $\kappa$
as a set of discrete grid cells or pixels $\kappa_i$ and rewrite the potential
(\eqnref{lensing potential}) as
%
\begin{equation}
  \psi(\vec\theta) = \sum_n \kappa_n Q_n(\vec\theta)
  \label{discrete potential}
\end{equation}
%
where the sum runs over all the pixels and $Q_n$ is the integral of the logarithm
over pixel $n$. The exact form for $Q$ is described in \appref{Q derivation}.
We can find the discretized lens equation by simply taking the gradient of the
above equations.

@@

\subsection{Raytracing}
\label{Raytracing}
\Glass\ can also determine the position of images and time delays from 
particle-based simulation output given a source position $\vec\beta$. This is
used to generate the lens configurations used in the parameter study.  The
particles are first projected onto a very high resolution grid representing the
lens plane. The centers $\vec\theta_i$ of each of the grid cells are mapped
back onto the source plane using \eqnref{eqn:lens_equation}. If the location on
the source plane $\vec\beta_i$ is within a user specified
$\eps_\mathrm{accept}$ of $\vec\beta$ then $\vec\theta_i$ is 
accepted and further refined using a root finding algorithm until the distance
to $\vec\beta$ is nearly zero. If multiple points converge to an
$\eps_\mathrm{root}$ of each other then only one point is taken.  Care must be
taken that the grid resolution is high enough that the resulting image position
error is below the equivalent observational error. Time delays are then
calculated in order from the arrival time at each image (\eqnref{tau}).

\section{The mock data}\label{sec:mockdata}

@@

\subsection{Setting up the triaxial N-body mocks}

@@

Each of these configurations were modelled with and without time delays and with
and without a central image. We assumed for all our tests that the lensing mass
was radially symmetric. By construction, this is known to be true and this is
most often the case with real galaxies, unless there is an obvious observed
asymmetry. The central pixel was refined into a further nine pixels to capture
any steep rise in the profile. Two of the four mock galaxies have a steeply rising
inner profile.

\section{Results}\label{sec:results}

@@

The main results from modelling these different configuration are shown black in
\figref{main results}. Each subplot corresponds to a different galaxy and
the vertical axis shows the range of quality.  The quality of a
model $\M_i$ by comparing the recovered density profile $\rho_\M(r)$ against
the profile of the mock galaxy $\rho_G(r)$.  In particular, the plots show the
RMS quantity
%
\begin{equation}
  \chi_i = 100 \times \sqrt{\frac{\int_r (\rho_{\M_i}(r) - \rho_G(r))^2}{\int_r \rho_G(r)^2}}
  %\chi_i = \frac{\sum_r (\rho(r)_{\M_i} - \rho(r)_G)^2}{\sum_r \rho(r)_G^2}
\end{equation}
%

@@

\section{Conclusions}\label{sec:conclusions}
